---
title: "First Principles Regression"
author: "Wyatt Clegg"
date: "December 6, 2016"
output: pdf_document
---


The first set of problems uses the data set `data1exam3.dat'.  Call the first column of the data y, the second column x1, and the third column x2. 
```{r}
setwd("~/Documents/School/BYU 2016-2017/Fall 2016/535/Final")
data1 <- read.table('data1exam3.dat')
colnames(data1)<-c("y","x1","x2")
firstprin<-function(y,X)
{
  n<-dim(X)[1]
  p<-dim(X)[2]
  xpx<-t(X)%*%X
  xpxi<-solve(xpx)
  bhat<-xpxi%*%t(X)%*%y
  yhat<-X%*%bhat
  RSS<-t(y-yhat)%*%(y-yhat)
  s2hat<-as.numeric(RSS)/(n-p)
  SYY<-as.numeric(t(y-mean(y))%*%(y-mean(y)))
  SSReg<-SYY - RSS
  R2 <- SSReg/SYY
  adjR2 <- X-s2hat/(SYY/(n-1))
  vhatbhat<- s2hat*xpxi
  tstar <-(bhat)/sqrt(s2hat/n)
  ei<-y-yhat
  hat<-X%*%xpxi%*%t(X)
  littleh<- diag(hat)
  stei<-ei/(sqrt(s2hat*(1-littleh)))
  stdel<-ei*sqrt((n-p-1)/(RSS*(1-littleh)-ei^2))
  dffits <- stdel*sqrt(littleh/(1-littleh))
  cooksd<- ((ei/(1-littleh))^2*littleh)/(p*s2hat)
  df<-n-p
  list(bhat=bhat,yhat=yhat,RSS=RSS,s2hat=s2hat,SYY=SYY,
       SSReg=SSReg,R2=R2,adjR2=adjR2,vhatbhat=vhatbhat,
       tstar=tstar,ei=ei,stei=stei,stdel=stdel,dffits=dffits,
       cooksd=cooksd,xpxi=xpxi,n=n,p=p,df=df) 
}

getF<-function(contrast,beta,vhatbhat){
  rp <- nrow(contrast)
  cc<- contrast
  b <- beta
  vh <-vhatbhat
  t(cc%*%b)%*%solve(cc%*%vh%*%t(cc))%*%(cc%*%b)/rp
}

getSS<-function(contrast,beta,xpxi){
  cc<-contrast
  b<-beta
  t(cc%*%b)%*%solve(cc%*%xpxi%*%t(cc))%*%(cc%*%beta)
}
```

1.  Using the model $E(y) = \beta_0 + \beta_{1} x1 + \beta_{2} x2$, find the $\hat\beta's$.
```{r}
x<-as.matrix(cbind(1,data1[,2:3]))
prob1<-firstprin(data1$y,x)
prob1$bhat

```
2. Produce a table of the $\hat\beta's$, the standard errors of the $\hat\beta's$, the t-statistics of the null hypotheses that the $\beta's$ are equal to 0, and the p-values for the (two-tailed) tests.
```{r}
SE<-qt(0.975,dim(x)[1]-dim(x)[2])*
  sqrt(diag(prob1$vhatbhat))

pvalues<-1-pt(prob1$bhat,dim(x)[1]-dim(x)[2])
cbind(prob1$bhat,SE,qt(0.975,dim(x)[1]-dim(x)[2]),pvalues)

```
3. Using the model in problem 1, find the F statistic to test the null hypothesis that $\beta_2=0$.
```{r}
C<-t(c(0,0,1))
Fstar<-getF(C,prob1$bhat,prob1$vhatbhat)
Fstar
```
4. Compute the F statistic to test the null hypothesis that $\beta_2=0$ using $F = \frac{TmtSS/Tmtdf}{ErrorSS/Errordf}$.
```{r}
Fstar2<-(getSS(C,prob1$bhat,prob1$xpxi)/1)/
  ((prob1$RSS)/(prob1$n-3))
Fstar2
```
5. Now fit the model $E(y) = \beta_0 + \beta_1 x1$.  What is $\hat\beta_1$?
```{r}
x2<-as.matrix(cbind(1,data1$x1))
fit2<-firstprin(data1$y,x2)
fit2$bhat[2]
```
6. Now fit the model $E(x_2) = \beta_0 + \beta_1 x1$.
What is $\hat\beta_1$?
```{r}
fit3<-firstprin(data1$x2,x2)
fit3$bhat[2]
```
7.  Take the residuals from the model in problem 5 and use them as the response, and fit them to a model using the residuals in problem 6 as the x's. That is, the model looks like $E(y-\hat{y}) = \beta_3 (x2-\hat{x2})$.  What is $\hat\beta_3$?
```{r}
e<-fit2$ei
x3<-fit3$ei
fit4<-firstprin(e,x3)
fit4$bhat
```
8.  How does $\hat\beta_3$ in problem 7 compare to $\hat\beta_2$ in problem 1?

They are identical.
```{r}
c(fit4$bhat,prob1$bhat[3])
```
9. Now consider an analysis of variance setting where there are 3 treatments each at 2 levels.  We will call them treatments A, B, and C.  So the cells could be labled A1B1C1, A1B1C2, A1B2C1, A1B2C2, A2B1C1, A2B1C2, A2B2C1, and A2B2C2.  The $\mu$ vector then would be the transpose of $(\mu_{111},\mu_{112},\mu_{121},\mu_{122},\mu_{211},\mu_{212},\mu_{221},\mu_{222})$.With 4 observations in each of the 8 cells, the data are found in the file 'data2exam3.dat'. What are the maximum likelihood estimates for the 8 cell means?
```{r}
data2 <- read.table('data2exam3.dat')
trt<-rep(1:8,each=4)
w<-model.matrix(~as.factor(trt)-1)
prob2<-firstprin(data2$y,w)
prob2$bhat
```

10.  What is the F statistic to test the null hypothesis that all the treatment mean parameters are equal?
```{r}
Q<-diag(1,7,7)
Q1<-cbind(Q,0)
C2<-Q1-cbind(0,Q)
Fstar3<-getF(C2,prob2$bhat,prob2$vhatbhat)
Fstar3
```
11.  What is the p-value of the F statistic in the previous problem?
```{r}
1-pf(Fstar3,7,prob2$df)
```
12.  Now I want you to use the following matrix to transform the $\mu's$ into $\beta's$.  
```{r}
a <- rbind(c(1/8,1/8,1/8,1/8,1/8,1/8,1/8,1/8),c(1/4,1/4,1/4,1/4,-1/4,-1/4,-1/4,-1/4),c(1/4,1/4,-1/4,-1/4,1/4,1/4,-1/4,-1/4),c(1,1,-1,-1,-1,-1,1,1),c(1/4,-1/4,1/4,-1/4,1/4,-1/4,1/4,-1/4),c(1,-1,1,-1,-1,1,-1,1),c(1,-1,-1,1,1,-1,-1,1),c(1,-1,-1,1,-1,1,1,-1))
a
```
The first row of $a$ would be used to compute an overall intercept or an estimate of the mean of all 8 $\mu's$.  The second row of $a$ is used to compute a $\beta$ that is the average of all the cell means at level 1 of treatment A minus the average of all the cell means at level 2 of treatment A. That is, it would be used to test the marginal effect of treatment A.

13. What would the fourth row of $a$ be used to test?

It tests the interaction between Treatment A and Treatment
B.

14. What would the fifth row of $a$ be used to test?

It tests the effect of Treatment C.

15. What would the sixth row of $a$ be used to test?

It tests the interaction between Treatment A and 
Treatment C.


16. What would the eighth row of $a$ be used to test?

Interaction between all three Treatments: Treatment A,
Treatment B, and Treatment C.

17.  What are the $\hat\beta's$?
```{r}
ai<-solve(a)
X3<-w%*%ai
prob17<-firstprin(data2$y,X3)
prob17$bhat[,1]
```
18. Print out $a\hat\mu$ and $\hat\beta$ side by side.
```{r}
cbind(a%*%prob2$bhat,prob17$bhat)
```
19. Print out $\hat\mu$ and $a^{-1}\hat\beta$ side by side.
```{r}
cbind(prob2$bhat,ai%*%prob17$bhat)
```
20. What is the F statistic to test the null hypothesis that there is no three way (ABC) interaction among the treatments?
```{r}
C4<-t(c(0,0,0,0,0,0,0,1))
Fstar4<-getF(C4,prob17$bhat,prob17$vhatbhat)
Ftobeat<-qf(0.95,1,prob17$df)
Fstar4
## Test whether significant
Fstar4>=Ftobeat
```
21. What is the F statistic to test the null hypothesis that there is no AC interaction among the treatments?
```{r}
C5<-t(c(0,0,0,0,0,1,0,0))
Fstar5<-getF(C5,prob17$bhat,prob17$vhatbhat)
Fstar5
## Test whether significant
Fstar5>=Ftobeat
```
22. Now take out all the observations in cell A2B2C2.  Now estimate the $\hat\mu's$.  Now there are only 7.
```{r}
data3 <- data2[-29:-32,]
```

23. Now estimate $\hat\beta's$ using the $a$ matrix as previously defined, but constrain the three way interaction to be 0.  What are the $\hat\beta's$? There will only be 7.
```{r}
tmt1<-rep(1:7,each=4)
w2<-model.matrix(~as.factor(tmt1)-1)
prob23<-firstprin(data3$y,w2)
prob23$bhat
```
24. Now estimate the $\hat\mu's$ using the $\hat\beta's$ you just estimated.  What are the new $\hat\mu's$?  There should be 8.

```{r}
w3<-cbind(w2,0)
X4<-w3%*%ai
X5<-X4[,-8]
prob24<-firstprin(data3$y,X5)
result24<-ai[,-8]%*%prob24$bhat
result24
```
25. What do you get when you multiply the eighth row of the $a$ matrix by the $\hat\mu's$?
```{r}
a[8,]%*%result24
```
 We get 0.
 
 
26. Explain why the answer in the previous problem makes sense.

 We constrained the interaction between Treatments A, B,
 and C to be 0 as part of back estimating. Therefore it is
 logical and expected that we should get 0 when we multiply
 our A matrix by our back-estimated values.
 
27. Print out the $\hat\mu's$ next to their standard errors.
```{r}
SE<-qt(0.975,dim(X5)[1]-dim(X5)[2])*sqrt(diag(prob24$vhatbhat))
SE<-c(SE,pi)
cbind(result24,SE)
```